A Plot is Worth a Thousand Tests: Assessing Residual Diagnostics with the Lineup Protocol

### 2023-04-03 paper.tex
copy in the appendix

>>>>

\appendix

\hypertarget{experiment-setup}{%
\section{Experiment setup}\label{experiment-setup}}

\hypertarget{mapping-of-subjects-to-experimental-factors}{%
\subsection{Mapping of subjects to experimental
factors}\label{mapping-of-subjects-to-experimental-factors}}

Mapping of subjects to experimental factors is an important part of
experiment design. Essentially, we want to maximum the difference in
factors exposed to a subject. For this purpose, we design an algorithm
to conduct subject allocation. Let \(L\) be a set of available lineups
and \(S\) be a set of available subjects. According to the experimental
design, the availability of a lineup is associated with the number of
subjects it can assign to. For lineups with uniform fitted value
distribution, this value is 11. And other lineups can be allocated to at
most five different subjects. The availability of a subject is
associated with the number of lineups that being allocated to this
subject. A subject can view at most 18 different lineups.

The algorithm starts from picking a random subject \(s \in S\) with the
minimum number of allocated lineups. It then tries to find a lineup
\(l \in L\) that can maximise the distance metric \(D\) and allocate it
to subject \(s\). Set \(L\) and \(S\) will be updated and the picking
process will be repeated until there is no available lineups or
subjects.

Let \(F_1,...,F_q\) be \(q\) experimental factors, and \(f_1, ...,f_q\)
be the corresponding factor values. We say \(f_i\) exists in \(L_{s}\)
if any lineup in \(L_{s}\) has this factor value. Similarly, \(f_if_j\)
exists in \(L_{s}\) if any lineup in \(L_{s}\) has this pair of factor
values. And \(f_if_jf_k\) exists in \(L_{s}\) if any lineup in \(L_{s}\)
has this trio of factor values. The distance metric \(D\) is defined
between a lineup \(l\) and a set of lineups \(L_{s}\) allocated to a
subject \(s\) if \(L_{s}\) is non-empty:

\footnotesize

\begin{equation*}
D =
C - \sum_{1\leq i \leq q}I(f_i\text{ exists in }L_{s}) - \sum_{\substack{1\leq i \leq q-1 \\ i < j \leq q}}I(f_if_j\text{ exists in }L_{s}) - \sum_{\substack{1\leq i \leq q - 2 \\ i < j \leq q - 1 \\ j < k \leq q}}I(f_if_jf_k\text{ exists in }L_{s})
\end{equation*}

\normalsize

where \(C\) is a sufficiently large constant such that \(D > 0\). If
\(L_{s}\) is empty, we define \(D = 0\).

The distance measures how different a lineup is from the set of lineups
allocated to the subject in terms of factor values. Thus, the algorithm
will try to allocate the most different lineup to a subject at each
step.

\hypertarget{data-collection-process}{%
\subsection{Data collection process}\label{data-collection-process}}

The survey data is collected via a self-hosted website designed by us.
The complete architecture is provided in Figure \ref{fig:tech}. The
website is built with the \texttt{Flask} \citep{flask} web framework and
hosted on \texttt{PythonAnywhere} \citep{pythonanywhere}. It is
configured to handle HTTP requests such that subjects can correctly
receive webpages and submit responses. Embedded in the resources sent to
subjects, the \texttt{jsPsych} front-end framework \citep{jspsych}
instructs subjects' browsers to render an environment for running
behavioral experiments. During the experiment, this framework will
automatically collect common behavioral data such as response time and
clicks on buttons. Subjects' responses are first validated by a
scheduled \texttt{Python} script run on the server, then push to a
Github repository. Lineup images shown to users are saved in multiple
Github repositories and hosted in corresponding Github pages. The URLs
to these images are resolved by \texttt{Flask} and bundled in HTML
files.

Once the participant is recruited from Prolific
\citep{palan2018prolific}, it will be redirected to the entry page of
our study website. An image of the entry page is provided in Figure
\ref{fig:entry-page}. Then, the participant needs to submit the online
consent form and fill in the demographic information as shown in
\ref{fig:consent-form} and \ref{fig:metadata} respectively. Before
evaluating lineups, participant also need to read the training page as
provide in Figure \ref{fig:training-page} to understand the process. An
example of the lineup page is given in Figure \ref{fig:lineup-page}. A
half of the page is taken by the lineup image to attract participant's
attention. The button to skip the selections for the current lineup is
intentionally put in the corner of the bounding box with smaller font
size, such that participants will not misuse this functionality.

{\centering \includegraphics[width=1\linewidth]{figures/experiment_tech}

\caption{Diagram of online experiment setup. The server-side of the study website uses Flask as backend hosted on PythonAnywhere. And the client-side uses jsPsych to run experiment.}\label{fig:tech}
\end{figure}

{\centering \includegraphics[width=1\linewidth]{figures/message}

\caption{The entry page of the study website.}\label{fig:entry-page}
\end{figure}

{\centering \includegraphics[width=1\linewidth]{figures/consent_form}

\caption{The consent form provided in the study website.}\label{fig:consent-form}
\end{figure}

{\centering \includegraphics[width=1\linewidth]{figures/metadata}

\caption{The form to provide demographic information.}\label{fig:metadata}
\end{figure}

{\centering \includegraphics[width=1\linewidth]{figures/training}

\caption{The training page of the study website.}\label{fig:training-page}
\end{figure}

{\centering \includegraphics[width=1\linewidth]{figures/lineup1}

\caption{The lineup page of the study website.}\label{fig:lineup-page}
\end{figure}

\hypertarget{demographics}{%
\section{Demographics}\label{demographics}}

Along with the responses to lineups, we have collected a series of
demographic information including age, pronoun, education background and
previous experience in studies involved data visualization. Table
\ref{tab:pronoun}, \ref{tab:age-group}, \ref{tab:education} and
\ref{tab:experience} provide summary of the demographic data.

It can be observed from the tables that most participants have Diploma
or Bachelor degrees, followed by High school or below and the survey
data is gender balanced. Majority of participants are between 18 to 39
years old and there are slightly more participants who do not have
previous experience than those who have.

\caption{\label{tab:pronoun}Summary of pronoun distribution of subjects recuritted in this study.}
\centering
\begin{tabular}[t]{lrrrrrrrr}
\toprule
Pronoun & Period I & \% & Period II & \% & Period III & \% & Total & \%\\
\midrule
He & 77 & 17.4 & 79 & 17.8 & 61 & 13.8 & 217 & 49.0\\
She & 78 & 17.6 & 77 & 17.4 & 61 & 13.8 & 216 & 48.8\\
Other & 5 & 1.1 & 4 & 0.9 & 1 & 0.2 & 10 & 2.3\\
 & 160 & 36.1 & 160 & 36.1 & 123 & 27.8 & 443 & 100.0\\
\bottomrule
\end{tabular}
\end{table}

\caption{\label{tab:age-group}Summary of age distribution of subjects recuritted in this study.}
\centering
\begin{tabular}[t]{lrrrr}
\toprule
Age group & Period I & Period II & Period III & Total\\
\midrule
18-24 & 83 & 86 & 51 & 220\\
25-39 & 69 & 63 & 63 & 195\\
40-54 & 6 & 8 & 6 & 20\\
55-64 & 2 & 3 & 3 & 8\\
\bottomrule
\end{tabular}
\end{table}

\caption{\label{tab:education}Summary of education distribution of subjects recuritted in this study.}
\centering
\begin{tabular}[t]{lllll}
\toprule
Education & Period I & Period II & Period III & Total\\
\midrule
High School or below & 41 (9.3\%) & 53 (12\%) & 33 (7.4\%) & 127 (28.7\%)\\
Diploma and Bachelor Degree & 92 (20.8\%) & 79 (17.8\%) & 66 (14.9\%) & 237 (53.5\%)\\
Honours Degree & 6 (1.4\%) & 15 (3.4\%) & 6 (1.4\%) & 27 (6.1\%)\\
Masters Degree & 21 (4.7\%) & 13 (2.9\%) & 16 (3.6\%) & 50 (11.3\%)\\
Doctoral Degree & 0 (0\%) & 0 (0\%) & 2 (0.5\%) & 2 (0.5\%)\\
\addlinespace
 & 160 (36.1\%) & 160 (36.1\%) & 123 (27.8\%) & 443 (100\%)\\
\bottomrule
\end{tabular}
\end{table}

\caption{\label{tab:experience}Summary of previous experience distribution of subjects recuritted in this study.}
\centering
\begin{tabular}[t]{lrrrr}
\toprule
Previous experience & Period I & Period II & Period III & Total\\
\midrule
No & 96 & 88 & 67 & 251\\
Yes & 64 & 72 & 56 & 192\\
\bottomrule
\end{tabular}
\end{table}

\hypertarget{effect-size-derivation}{%
\section{Effect size derivation}\label{effect-size-derivation}}

Effect size can be defined as the difference of a parameter for a
particular model or distribution, or a statistic derived from a sample.
Importantly, it needs to reflect the treatment we try to measure.
Centred on a conventional statistical test, we usually can deduce the
effect size from the test statistic by substituting the null parameter
value. When considering the diagnostics of residual departures, there
exist many possibilities of test statistics for a variety of model
assumptions. Meanwhile, diagnostic plots such as the residual plot have
no general agreement on measuring how strong a model violation pattern
is. To build a bridge between various residual-based tests, and the
visual test, we focus on the shared information embedded in the testing
procedures, which is the distribution of residuals. When comes to
comparison of distribution, Kullback-Leibler divergence
\citep{kullback1951information} is a classical way to represent the
information loss or entropy increase caused by the approximation to the
true distribution, which in our case, the inefficiency due to the use of
false model assumptions.

Following the terminology introduced by \citet{kullback1951information},
\(P\) represents the measured probability distribution, and \(Q\)
represents the assumed probability distribution. The Kullback-Leibler
divergence is defined as
\(\int_{-\infty}^{\infty}log(p(x)/q(x))p(x)dx\), where \(p(.)\) and
\(q(.)\) denote probability densities of \(P\) and \(Q\).

Let \(\boldsymbol{X}_a = (\boldsymbol{1}, \boldsymbol{X})\) denotes the
\(p\) regressors with \(n\) observations,
\(\boldsymbol{R}_a = \boldsymbol{I} -\boldsymbol{X}(\boldsymbol{X}'\boldsymbol{X})^{-1}\boldsymbol{X}'\)
denotes the residual operator, and let
\(\boldsymbol{\varepsilon} \sim N(\boldsymbol{0},\sigma^2\boldsymbol{I})\)
denotes the error. Using the Frisch--Waugh--Lovell theorem, residuals
\(\boldsymbol{e} = \boldsymbol{R}_a\boldsymbol{\varepsilon}\). Because
\(rank(\boldsymbol{R}_a) = n - p < n\), \(e\) follows a degenerate
multivariate normal distribution and does not have a density. Since the
Kullback-Leibler divergence requires a proper density function, we need
to simplify the covariance matrix of \(\boldsymbol{e}\) by setting all
the off-diagonal elements to 0. Then, the residuals will assumed to
follow \(N(\boldsymbol{0}, diag(\boldsymbol{R}_a\sigma^2))\) under the
null hypothesis that the model is correctly specified. If the model is
however misspecified due to omitted variables \(\boldsymbol{Z}\), or a
non-constant variance \(\boldsymbol{V}\), the distribution of residuals
can be derived as
\(N(\boldsymbol{R}_a\boldsymbol{Z}\boldsymbol{\beta}_z, diag(\boldsymbol{R}_a\sigma^2))\)
and
\(N(\boldsymbol{0}, diag(\boldsymbol{R}_a\boldsymbol{V}\boldsymbol{R}_a'))\)
respectively.

By assuming both \(P\) and \(Q\) are multivariate normal density
functions, the Kullback-Leibler divergence can be rewritten as
\[KL = \frac{1}{2}\left(log\frac{|\Sigma_p|}{|\Sigma_q|} - n + tr(\Sigma_p^{-1}\Sigma_q) + (\mu_p - \mu_q)'\Sigma_p^{-1}(\mu_p - \mu_q)\right).\]

Then, we can combine the two residual departures into one formula

\small

\begin{equation}
\label{eq:effect-size}
KL = \frac{1}{2}\left(log\frac{|diag(\boldsymbol{R}_a\boldsymbol{V}\boldsymbol{R}_a')|}{|diag(\boldsymbol{R}_a\sigma^2)|} - n + tr(diag(\boldsymbol{R}_a\boldsymbol{V}\boldsymbol{R}_a')^{-1}diag(\boldsymbol{R}_a\sigma^2)) + \boldsymbol{\mu}_z^{T}(\boldsymbol{R}_a\boldsymbol{V}\boldsymbol{R}_a')^{-1}\boldsymbol{\mu}_z\right).
\end{equation}

\normalsize

When there are omitted variables but constant error variance, the
formula can be reduced to

\[KL = \frac{1}{2}\left(\boldsymbol{\mu}_z^{T}(diag(\boldsymbol{R}_a\sigma^2))^{-1}\boldsymbol{\mu}_z\right).\]

And when the model equation is correctly specified but the error
variance is non-constant, the formula can be reduced to

\[KL = \frac{1}{2}\left(log\frac{|diag(\boldsymbol{R}_a\boldsymbol{V}\boldsymbol{R}_a')|}{|diag(\boldsymbol{R}_a\sigma^2)|} - n + tr(diag(\boldsymbol{R}_a\boldsymbol{V}\boldsymbol{R}_a')^{-1}diag(\boldsymbol{R}_a\sigma^2))\right).\]
Since we assume \(\sigma = 1\) for the heteroskedasticity model, the
final form of the formula is

\[KL = \frac{1}{2}\left(log\frac{|diag(\boldsymbol{R}_a\boldsymbol{V}\boldsymbol{R}_a')|}{|diag(\boldsymbol{R}_a)|} - n + tr(diag(\boldsymbol{R}_a\boldsymbol{V}\boldsymbol{R}_a')^{-1}diag(\boldsymbol{R}_a))\right).\]

\hypertarget{effect-of-data-collection-period}{%
\section{Effect of data collection
period}\label{effect-of-data-collection-period}}

{\centering \includegraphics[width=1\linewidth]{paper_files/figure-latex/poly-boxplot-lineup-1}

\caption{A lineup of "letter-value" boxplots of weighted propotion of detect for lineups over different data collection periods for non-linearity model. Can you find the most different boxplot? The data plot is positioned in panel $2^3 - 1$.}\label{fig:poly-boxplot-lineup}
\end{figure}

{\centering \includegraphics[width=1\linewidth]{paper_files/figure-latex/heter-boxplot-lineup-1}

\caption{A lineup of "letter-value" boxplots of weighted propotion of detect for lineups over different data collection periods for heteroskedasticity model. Can you find the most different boxplot? The data plot is positioned in panel $2^4 - 2$.}\label{fig:heter-boxplot-lineup}
\end{figure}

We have the same type of model collected over different data collection
periods, that may lead to unexpected batch effect. Figure
\ref{fig:poly-boxplot-lineup} and \ref{fig:heter-boxplot-lineup} provide
two lineups to examine whether there is an actual difference across data
collection periods for non-linearity model and heteroskedasticity model
respectively. To emphasize the tail behaviour and display fewer
outliers, we use the ``letter-value'' boxplot \citep{hofmann2017value}
which is an extension of the number of ``letter value'' statistics to
check the weighed proportion of detect over different data collection
period. The weighted proportion of detect is calculated by taking the
average of \(c_i\) of a lineup over a data collection period. Within our
research team, we can not identify the data plot from the null plots for
these two lineups, result in \(p\)-values much greater than \(5\)\%.
Thus, there is no clear evidence of batch effect.

\hypertarget{sensitivity-analysis-for-alpha}{%
\section{\texorpdfstring{Sensitivity analysis for
\(\alpha\)}{Sensitivity analysis for \textbackslash alpha}}\label{sensitivity-analysis-for-alpha}}

The parameter \(\alpha\) used for the \(p\)-value calculation needs to
be estimated from responses to null lineups. With a greater value of
\(\hat{\alpha}\), the \(p\)-value will be smaller, resulting in more
lineups being rejected. However, The way we generate Rorschach lineup is
not strictly the same as what suggested in
\citet{vanderplas2021statistical} and \citet{buja_statistical_2009}.
Therefore, we conduct a sensitivity analysis in this section to examine
the impact of the variation of the estimator \(\alpha\) on our primary
findings.

The analysis is conducted by setting up several scenarios, where the
\(\alpha\) is under or overestimated by 12.5\%, 25\% and 50\%. Using the
adjusted \(\hat{\alpha}\), we recalculate the \(p\)-value for every
lineup and show the results in Figure \ref{fig:sensitivity}. It can be
observed that there are some changes to \(p\)-values, especially when
the \(\hat{\alpha}\) is multiplied by 50\%. However, Table
\ref{tab:sensitivity} shows that adjusting \(\hat{\alpha}\) will not
result in a huge difference in rejection decisions. There are only a
small percentage of cases where the rejection decision change. It is
very unlikely the downstream findings will be affected because of the
estimate of \(\alpha\).

{\centering \includegraphics[width=1\linewidth]{paper_files/figure-latex/sensitivity-1}

\caption{Change of $p$-values with $\hat{\alpha}$ multiplied by $0.5$, $0.75$, $0.875$, $1.125$, $1.25$ and $1.5$. The vertical dashed line is to indicate $p\text{-value} = 0.05$. The x-axis is drawn on logarithmic scale. For multiplier smaller than $1$, the adjusted $p$-value will incrase then decrase as $p$-value increases. The trend is the opposite for multiplier greater than $1$, but the difference will eventually reach $0$.}\label{fig:sensitivity}
\end{figure}

\caption{\label{tab:sensitivity}Change of rejection decision because of the modification of $\hat{\alpha}$.}
\centering
\begin{tabular}[t]{rll}
\toprule
multiplier & Proportion of lineups transforms to "not reject" & Proportion of lineups transforms to "reject"\\
\midrule
0.500 & 2.51\% & 0\%\\
0.750 & 1.43\% & 0\%\\
0.875 & 1.08\% & 0\%\\
1.125 & 0\% & 1.08\%\\
1.250 & 0\% & 1.43\%\\
\addlinespace
1.500 & 0\% & 1.79\%\\
\bottomrule
\end{tabular}
\end{table}

\hypertarget{effect-of-number-of-evaluations-on-the-power-of-a-visual-test}{%
\section{Effect of number of evaluations on the power of a visual
test}\label{effect-of-number-of-evaluations-on-the-power-of-a-visual-test}}

When comparing power of visual tests across different fitted value
distributions, we have discussed the number of evaluations on a lineup
will affect the power of the visual test. Using the lineups with uniform
fitted value distribution, we show in Figure \ref{fig:n_eval} the change
of power of visual tests due to different number of evaluations. It can
be learned that as the number of evaluations increases, the power will
increase but the margin will decrease. Considering we have eleven
evaluations on lineups with uniform fitted value distribution, and five
evaluations on other lineups, it is necessary to use the same number of
evaluations for each lineup in comparison.

{\centering \includegraphics[width=1\linewidth]{paper_files/figure-latex/n_eval-1}

\caption{Change of power of visual tests for different number of evalutions on lineups with uniform fitted value distribution. The power will incrase as the number of evaluations increases, but the margin will decrase.}\label{fig:n_eval}
\end{figure}

\hypertarget{power-of-a-reset-test-under-different-auxiliary-regression-formulas}{%
\section{Power of a RESET test under different auxiliary regression
formulas}\label{power-of-a-reset-test-under-different-auxiliary-regression-formulas}}

It is found in the result that the power of a RESET test will be
affected by the highest order of fitted values included in the auxiliary
formula. And we suspect that the current recommendation of the highest
order - four, is insufficient to test complex non-linear structures such
as the ``Triple-U'' shape designed in this paper. Figure \ref{fig:reset}
illustrates the change of power of RESET test while testing the ``U''
shape and the ``Triple-U'' shape with different highest orders. Clearly,
when testing a simple shape like the ``U'' shape, the highest order has
very little impact on the power. But for testing the ``Triple-U'' shape,
there will be a loss of power if the recommended order is used. To avoid
the loss of power, the highest order needs to be set to at least six.

{\centering \includegraphics[width=1\linewidth]{paper_files/figure-latex/reset-1}

\caption{Change of power of RESET tests for different orders of fitted values included in the auxiliary formula. The left panel is the power of testing the "U" shape and the right panel is the power of testing the "Triple-U" shape. The power will not be greatly affected by the highest order in the case of testing the "U" shape. In the case of testing the "Triple-U" shape, the highest order needs to be set to at least six to avoid the loss of power.}\label{fig:reset}
\end{figure}

\clearpage
